A fast primal-dual algorithm for computing Nash equilibria and best response strategies in two-person zero-sum sequential games

### 2014-08-08 paper.tex
yet another endeavor towards game tree abstraction

\documentclass[a4paper,10pt,journal]{IEEEtran}
\usepackage{graphicx} \usepackage[cmex10]{amsmath}
\usepackage{amssymb} \usepackage{fancybox} \usepackage{alltt}
\usepackage{soul} \usepackage{color} \usepackage{verbatim}
\usepackage{xcolor} \usepackage{colortbl,hhline}
\usepackage[ruled,vlined]{algorithm2e} \usepackage{framed}
\usepackage{amsthm} \usepackage{fancyref}
\usepackage{amsmath,amsfonts,amssymb, amsthm} \usepackage{textcomp}

@@

\begin{document}
%\onecolumn
\title{A fast primal-dual algorithm for computing best response strategies in two-person sequential games}

@@

\begin{abstract}
% LESS THAN 200 WORDS !!!!!
% 
% While medical imaging typically provides massive amounts of
% data, the extraction of relevant information in a given
% applicative context remains a difficult challenge.
% XXX : don't call it SPM but just "predictive regions or predictive
% maps"
In this manuscript, we consider the problem of computing a best response against an opponent's realization plan in two-person sequential games.
The proposed algorithm for solving the corresponding contrained convex-optimization problem, derives from the primal-dual scheme of A. Chambolle and T. Pock.
Our algorithm is simple: all resolvent operators can be effectively computed in closed-form, using only elementary algebraic operations.
%%  and is orders of magnitude more efficient than state-of-the-art algorithms like those based on Linear Programming, Interior-Point methods,
%% and more recently, the Nesterov Excessive Gap Technique.
By way of example, we conclude by exhibiting results on Texas Hold'em Poker.
\end{abstract}

\begin{IEEEkeywords}
  game theory; two-person sequential game; best response; convex-optimization; contraint; primal-dual scheme
\end{IEEEkeywords}

\section{Introduction}
%\label{sec:intro}
\subsection{Notation and Terminology}
To begin, let us introduce some technical notation and terminology we will be using in this paper. The reader should lookup any standard textbook
(for example \cite{boyd2004}) on convex optimization for a tutorial introduction to these notions. Viz,
\begin{itemize}
\item $A^T$: \quad transpose of the matrix $A$;
\item $\begin{bmatrix}A\\B\end{bmatrix}$: vertical stacking of the matrices $A$ and $B$;
\item $v_j$: \quad $j$th component of the vector $v$;
\item $u^Tv$: \quad the dot product of two vectors $u$ and $v$;
\item $\mathbb{R}_+$: \quad the nonnegative part of the real-line $\mathbb{R}$;
\item $(x)_+$: \quad the component-wise maximum of the vector $x$ and 0;
\item $i_C$: \quad the indicator function of the convex set $C$;
\item $\Pi_C$: \quad the euclidean projector onto the convex set $C$;
\item $\|K\|_2$: \quad the matrix 2-norm of the matrix $K$;
\item $f^*$: \quad the Fenchel transform of the convex l.s.c function $f$;
\item \textit{l.s.c}: \quad acronyme for ``lower semi-continuous'';
\item $E^{-1}(U)$: \quad the pre-image of $U \subset \mathbb{R}^n$ under a matrix $E \in \mathbb{R}^{p,n}$,
namely the set $\{x \in \mathbb{R}^n\text{ }| \text{ }Ex \in U\}$
\end{itemize}

\subsection{Statement of the problem}
Consider a two-person sequential game\footnote{In a sequential game, players take turns in play, one after the other,
as opposed to simultaneous play.} in \textit{sequential-form} (See for example, \cite{von1996efficient} and \cite{koller1996efficient} 
for theory on sequential-form representation), and let $A \in \mathbb{R}^{m,n}$ be our payoff matrix.
We will be referring to the other player as ``the opponent''. We are interested in the problem of
finding a best response strategy $x^* \in \mathbb{R}_{+}^n \cap E^{-1}(\{e\})$, given a fixed
(possibly mixed) strategy $y_0$ for the opponent, where $E$ is a sparse $p$-by-$n$ matrix whose entries are $-1$, $0$, or $+1$,
and $(1, 0, 0, ..., 0) =: e \in \mathbb{R}^p$. Recall that $E$ and $e$ encode linear constraints on our ``admissible''
realization plans\footnote{In games in sequential form, the terms ``strategy'' and ``realization plan''
mean thesame thing.} $x$, in the sequential form representation of the game.
In the language of convex-optimization,
one can readily give the following saddle-point formulation for this problem. Viz,

@@

\begin{remark}
  In practice, $A$ and $E$ are very sparse as a game will typically have very few leafs
(i.e terminal states). This sparsity should be thoroughly exploited by a solver for problem \eqref{eq:opt_pb}.
\end{remark}

@@

\section{The proposed algorithm}
\label{sec:algo}
In this section we present the algorithm which is the purpose of this paper, namely an algorithm
for solving \eqref{eq:opt_pb}. Our algorithm (Alg.\ref{Tab:algo}) is a use-case of
the generic primal-dual algorithm of A. Chambolle and T. Pock, namely Algorithm 1 of \cite{chambolle2010}.

@@

Furthermore, one easily checks that $F^*$ and $G$ are \textit{proper} convex l.s.c, with resolvent
(a.k.a \textit{proximity}) operators given by the simple formulae:
\begin{equation}
  \left.
  \begin{aligned}
    (1 + \tau \partial G)^{-1} = \Pi_{\mathbb{R}_+^n}: x \mapsto (x)_+\\
    (1 + \sigma \partial F^*)^{-1}: (y, \zeta) \mapsto (y_0, \zeta + \sigma e)
  \end{aligned}
  \right\}
\end{equation}

@@

\begin{algorithm}[htb]
  \caption{Primal-dual algorithm for computing best response against opponent's realization plan $y_0$}
  \textbf{Given} Tolerance $\epsilon > 0$.\\
  \textbf{Initialize} $\tilde{x^{(0)}} = x^{(0)} \in \mathbb{R}^n$; $\zeta^{(0)} \in \mathbb{R}^{q}$;
  $\tau, \sigma > 0 \text{ s.t. }\tau\sigma \|K\|_2^2 < 1$; $k = 0$.\\
  \textbf{Precompute} $\eta_0 \leftarrow \tau A^Ty_0$\\
  \Repeat{
    $\frac{\|x^{(k+1)} - x^{(k)}\|^2_2}{2\tau} + \frac{\|\zeta^{(k+1)} - \zeta^{(k)}\|^2_2}{2\sigma} < \epsilon$}{
    \begin{eqnarray*}
      \zeta^{(k+1)} &\leftarrow& \zeta^{(k)} - \sigma \bigl(e - E\tilde{x}^{(k)}\bigr)\\
      x^{(k+1)} &\leftarrow& \bigl(x^{(k)} + \eta_0 + E^T\zeta^{(k+1)}\bigr)_+\\
      \tilde{x}^{(k+1)} &\leftarrow& 2x^{(k+1)} - x^{(k)}\\
      k &\leftarrow& k + 1
    \end{eqnarray*}
  } \Return $x^{(k)}$, $-y_0^TAx^{(k)}$
  \label{Tab:algo}
\end{algorithm}

>>>>

\documentclass[a4paper,9pt,journal]{IEEEtran}
\usepackage{graphicx} \usepackage[cmex10]{amsmath}
\usepackage{amssymb} \usepackage{fancybox} \usepackage{alltt}
\usepackage{soul} \usepackage{color} \usepackage{verbatim}
\usepackage{xcolor} \usepackage{colortbl,hhline}
\usepackage[ruled,vlined]{algorithm2e} \usepackage{framed}
\usepackage{amsthm} \usepackage{fancyref}
\usepackage{amsmath,amsfonts,amssymb, amsthm} \usepackage{textcomp}

@@

\begin{document}
%\onecolumn
\title{A fast primal-dual algorithm for computing Nash equilibria and best response strategies in two-person zero-sum sequential games}

@@

\begin{abstract}
% LESS THAN 200 WORDS !!!!!
% 
% While medical imaging typically provides massive amounts of
% data, the extraction of relevant information in a given
% applicative context remains a difficult challenge.
% XXX : don't call it SPM but just "predictive regions or predictive
% maps"
In this manuscript, we consider the problem of computing a best response against an opponent's realization plan in two-person sequential games.
The proposed algorithm for solving the corresponding contrained convex-optimization problem, derives from the primal-dual scheme of A. Chambolle and T. Pock.
Our algorithm is simple: all resolvent operators can be effectively computed in closed-form, using only elementary algebraic operations.
%%  and is orders of magnitude more efficient than state-of-the-art algorithms like those based on Linear Programming, Interior-Point methods,
%% and more recently, the Nesterov Excessive Gap Technique.
% By way of example, we conclude by exhibiting results on Texas Hold'em Poker.
\end{abstract}

\begin{IEEEkeywords}
  game theory; two-person sequential game; Nash equilibria; best response; convex-optimization; contraint; primal-dual scheme
\end{IEEEkeywords}

\section{Introduction}
%\label{sec:intro}
\subsection{Notation and Terminology}
To begin, let us introduce some technical notation and terminology we will be using in this paper. The reader should lookup any standard textbook
(for example \cite{boyd2004}) on convex optimization for a tutorial introduction to these notions. Viz,
\begin{itemize}
\item $\mathbb{R}^{m, n}$ \quad space of all $m$-by-$p$ real matrices;
\item $E^{-1}(U)$: \quad pre-image of $U \subset \mathbb{R}^n$ under a matrix $E \in \mathbb{R}^{p,n}$,
namely the set $\{x \in \mathbb{R}^n\text{ }| \text{ }Ex \in U\}$
\item $A^T$: \quad transpose of a matrix $A$;
\item $\begin{bmatrix}A\\B\end{bmatrix}$: \quad vertical stacking of two matrices $A$ and $B$;
\item $v_j$: \quad $j$th component of  a vector $v$;
\item $u^Tv$: \quad dot product of two vectors $u$ and $v$;
\item $\mathbb{R}^n_+$: \quad the $n$-dimensional nonnegative orthant;
\item $(x)_+$: \quad component-wise maximum of a vector $x$ and 0;
\item $i_C$: \quad indicator function of a convex set $C$;
\item $\Pi_C$: \quad euclidean projector onto a convex set $C$;
\item $\|K\|_2$: \quad matrix 2-norm of a matrix $K$;
\item \textit{p.c.l.s.c}: \quad acronym for \textit{proper convex lower semi-continuous};
\item $f^*$: \quad Fenchel transform (a.k.a convex conjugate) of a \textit{p.c.l.s.c} function $f$;
\item $(1 + \sigma \partial f)^{-1}$: resolvent (a.k.a proximal) operator of a \textit{p.c.l.s.c} function $f$, for a given stepsize $\sigma > 0$
\end{itemize}

\subsection{Statement of the problem}
Consider a two-person sequential game\footnote{In a sequential game, players take turns in play, one after the other,
as opposed to simultaneous play.} in \textit{sequential-form} (See for example, \cite{von1996efficient}
and \cite{koller1996efficient} 
for theory on sequential-form representation), and let $A \in \mathbb{R}^{m,n}$ be our payoff matrix.
We will be referring to the other player as ``the opponent''. We are interested in the problem of
finding a best response strategy $x^* \in \mathbb{R}_{+}^n \cap E^{-1}(\{e\})$, given a fixed
behavioral strategy $y_0$ for the opponent, where $E$ is a sparse $p$-by-$n$ matrix whose
entries are $-1$, $0$, or $+1$, and $(1, 0, 0, ..., 0) =: e \in \mathbb{R}^p$. Recall that $E$
and $e$ encode linear constraints on our ``admissible'' realization plans\footnote{In games
in sequential form, the terms ``strategy'' and ``realization plan'' mean thesame thing.} $x$,
in the sequential form representation of the game. In the language of convex-optimization,
one can readily give the following saddle-point formulation for this problem. Viz,

@@

\begin{remark}
  In practice, $A$, $E$, and $F$ are very sparse: $A$ will be sparse because a concrete sequential game will
typically have very few\footnote{Few, relative to the size of the game tree.} leafs, and only a few
combinations of sequences of moves of the players, will actually lead to a leaf (i.e. end the game);
$E$ and $F$ will be sparse because the kinks of possible sequences of moves of each player will
zig-zag between only a limited number of the player's information sets so that a move at an information set will
rarely\footnote{Relative to the number of information sets for the player.}  extend another information set.
This sparsity should be thoroughly exploited\footnote{For example when
multiplying vectors with these matrices.} by a solver for problem \eqref{eq:opt_pb}.
\end{remark}

@@

\section{The proposed algorithm}
0\label{sec:algo}
In this section we present the algorithm which is the purpose of this paper, namely an algorithm
for solving \eqref{eq:opt_pb}. Our algorithm (Alg.\ref{Tab:algo}) is a use-case of
the generic primal-dual algorithm of A. Chambolle and T. Pock, namely Algorithm 1 of \cite{chambolle2010}.

@@

Furthermore, one easily checks that $F^*$ and $G$ are \textit{p.c.l.s.c} with resolvent
(a.k.a \textit{proximity}) operators given by the simple formulae:
\begin{equation}
  \left.
  \begin{aligned}
    (1 + \tau \partial G)^{-1} = \Pi_{\mathbb{R}_+^n}: x \mapsto (x)_+\\
    (1 + \sigma \partial F^*)^{-1}: (y, \zeta) \mapsto (y_0, \zeta + \sigma e)
  \end{aligned}
  \right\}
\end{equation}

@@

\begin{algorithm}[htb]
  \caption{Primal-dual algorithm for computing best response against opponent's fixed realization plan $y_0$}
  \textbf{Given} Tolerance $\epsilon > 0$.\\
  \textbf{Initialize} $\tilde{x^{(0)}} = x^{(0)} \in \mathbb{R}^n$; $\zeta^{(0)} \in \mathbb{R}^{q}$;
  $\tau, \sigma > 0 \text{ s.t. }\tau\sigma \|K\|_2^2 < 1$; $k = 0$.\\
  \textbf{Precompute} $\eta_0 \leftarrow \tau A^Ty_0$\\
  \Repeat{
    $\frac{\|x^{(k+1)} - x^{(k)}\|^2_2}{2\tau} + \frac{\|\zeta^{(k+1)} - \zeta^{(k)}\|^2_2}{2\sigma} < \epsilon$}{
    \begin{eqnarray*}
      \zeta^{(k+1)} &\leftarrow& \zeta^{(k)} - \sigma \bigl(e - E\tilde{x}^{(k)}\bigr)\\
      x^{(k+1)} &\leftarrow& \bigl(x^{(k)} + \eta_0 + E^T\zeta^{(k+1)}\bigr)_+\\
      \tilde{x}^{(k+1)} &\leftarrow& 2x^{(k+1)} - x^{(k)}\\
      k &\leftarrow& k + 1
    \end{eqnarray*}
  } \Return $x^{(k)}$, $-y_0^TAx^{(k)}$
  \label{Tab:algo}
\end{algorithm}
